---
title: "Real Data Analysis For Methylation Experiments"
author: "Kushal Kumar Dey"
date: "Sunday, March 08, 2015"
output: html_document
---

In this script, we shall apply the probe level Methylation Topic Model on a real dataset. The dataset I consider in this example is the GSE52865 dataset in the GEO Omnibus, which crresponds to the paper **A DNA methylation-based definition of biologically distinct breast cancer subtypes**. In this paper, the authors have tried to address various types and sub-types of cancer through heat maps and hierarchical agglomerative clustering. We shall employ a topic modeling tpe algorithm to see how model based admixture type clustering would work under a situtation like this. We select the 2000 top CpG sites having the highest coefficient of variation (CV). 

The dataset is stored in the folder *Methylation* under the home directory *kkdey* in the RCC Midway Cluster. I produce two text files, one with the beta intensity expression values across all the samples for the top varying $1000$ CG sites and the other, for the top varying $10,000$ CG sites. The files were named **betas_file_1000.txt** and **betas_file_10000_txt** and in the following R chunk I obtain the bivariate plots of First and Second PC, and also First PC and Third PC to see how the normal samples differ from tumor samples. 


```{r,echo=TRUE,eval=TRUE}

#############  Exploration of the Data for the top 10,000 CpG sites selected using highest CV   #######################################

methData=read.table("D:/Matthew_Stephens_Project/Methylation/Methylation Expt 2/betas_file_10000.txt");
betas=t(methData)

###  now run the remaining part of the Methyl.estimation.R file 

###  The number of clusters in the real Data -  by Principal Component Analysis

###  Probably from the PC plots (PC1 vs PC2) or the Circular Projection plot, we

###   can get some intuitive idea


Princomp.analysis=prcomp(betas,center=TRUE);
eigenPrincomp=Princomp.analysis$sdev;
rotPrincomp=Princomp.analysis$rotation;
windows()
barplot(eigenPrincomp,col="red", 
  	ylab="eigenvalues of beta value matrix",
		xlab="Labels of clusters",xaxt="n");
#axis(1,at=0.7+1.2*(0:6),c("PC1","PC2","PC3","PC4","PC5","PC6","PC7"));

projection=t(rotPrincomp[,-dim(rotPrincomp)[2]])%*%t(betas);

windows()
plot(projection[1,],projection[2,],main="FirstPC vs SecondPC plot",xlab="First PC",
	ylab="Second PC",col=c(rep("red",17),rep("blue",40)));
legend("bottomright",c("N samples", "T samples"),fill=c("red","blue"));

windows()
plot(projection[1,],projection[3,],main="FirstPC vs ThirdPC plot",xlab="First PC",
	ylab="Third PC",col=c(rep("red",17),rep("blue",40)));
legend("bottomright",c("N samples", "T samples"),fill=c("red","blue"));

methData_reduced=methData[1:2000,];


```

Now we apply a topic model type approach on the best $1000$ CpG sites. so, we consider $1000 \times 57$ matrix and try to fit a Beta Topic model to that data.

```{r,eval=TRUE,echo=TRUE}
  reverse_transform=function(x) 
  {
    out=array(0,K-1);
		for(i in 1:(K-1))
		{
			out[i]=log((x[i]+10^(-5))/(x[K]+10^(-5)));
		}
		return(out)
	}

# Data
  

	# Log-Likelihood
	loglik_norm <- function(u,y,x) sum((y - x %*% u)^2)

	# Transform the parameters: we just have
	# to find a bijection between R^3 
	# and {(a,b,c,d) \in [0,1]^4 : a+b+c+d=1}.

  transform <- function(v) 
  {
  	# Ensure they in [0,1]
  	temp =c(exp(v),1);
  	out=temp/(1+sum(exp(v)));
  	return(out)
	}



```

Now we consider the **methData_reduced** table that I obtained at the end of the initial R chunk which has $2000$ rows corresponding to the top varying CpG sites and $57$ columns corresponding to the tissue samples of which the first $17$ correspond to the normal tissue samples and the last $40$ corresponds to the tumor tissue samples. 


```{r,echo=TRUE,eval=TRUE}


meth=t(methData_reduced);

scale=1; N=dim(meth)[1]; G=dim(meth)[2];

K=6;

####  The starting value of omega (topic prop weights)  ####################

# Use a preset seed so the example is reproducable.
require("setRNG")
old.seed <- setRNG(list(kind="Mersenne-Twister", normal.kind="Inversion",
seed=54321))
library(gtools)
#omega_preprocess=matrix(rdirichlet(N,c(scale/K,scale/K,scale/K,scale/K)), nrow=N);
omega_preprocess=matrix(rdirichlet(N,rep(scale/K,K)), nrow=N);



logit_meth=log(meth+1e-07)-log(1-meth+1e-07);
y=matrix(logit_meth,1,N*G);


omega0=omega_preprocess
alpha0=matrix(rnorm((K)*G,1,1),nrow=(K));
param0=c(matrix(omega0,1,N*K),matrix(alpha0,1,K*G));

library(SQUAREM)

beta_process.probe <- function(param_vec,y)
{
  logit_meth=matrix(y,N,G);
  #N=dim(logit_meth)[1]; G=dim(logit_meth)[2];
	omega0=matrix(param_vec[1:(N*K)],N,K);
	alpha0=matrix(param_vec[-(1:(N*K))],K,G);
	
	svd_omega=svd(omega0);
	temp1=t(svd_omega$v)%*%diag(1/svd_omega$d^2,dim(omega0)[2])%*%svd_omega$v;
	temp2=t(omega0)%*%logit_meth;
	# temp2=t(omega0)%*%pnorm_meth;
	temp1=solve(t(omega0)%*%omega0);
	H = temp1%*%temp2;

	###  Estimation of the matrix W (or omega) 

	omega=matrix(0,N,K);
	for(n in 1:N)
	{
		omega_vec=omega0[n,];
		meth_vec=logit_meth[n,];
		res=optim(reverse_transform(omega_vec), function(v) loglik_norm(transform(v),meth_vec,t(H)) );
		omega[n,]=transform(res$par);
	}

	param_vec_omega=matrix(omega,1,N*K);
	param_vec_alpha=matrix(H,1,K*G);
	param_new_vec=c(param_vec_omega,param_vec_alpha);
	param_vec=param_new_vec
	return(param_new_vec)
}

options(warn=-1)
system.time(res <- squarem(p=param0,y=y, fixptfn=beta_process.probe, control=list(maxiter = 100, trace = FALSE)));

omega_initial=matrix(res$par[1:(N*K)],N,K);
alpha_initial=matrix(res$par[-(1:(N*K))],K,G)

# docweights=omega_initial;
# library(permute);
# library("BioPhysConnectoR");
# perm_set=rbind(1:K,allPerms(1:K));
# diff=array(0,dim(perm_set)[1]);
# for (p in 1:dim(perm_set)[1])
# {
#  temp=docweights[,perm_set[p,]];
#	diff[p]=fnorm(temp,omega_true);
# }

# p_star=which(diff==min(diff));
# docweights=docweights[,perm_set[p_star,]];

# omega_initial=docweights;
# alpha_initial=alpha_initial[perm_set[p_star,],];

windows()
k=K
barplot(t(omega_initial),col=2:(k+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of Preprocessing topic proportions,k=",K))

```

The final iterative algorithm for estimating the topic proportions and the effect sizes of the beta Topic model is given as follows

```{r,echo=TRUE,eval=TRUE}

meth=t(methData_reduced);

y=matrix(meth,1,N*G);


omega_loglik_beta = function(omega_vec,meth_vec,alpha,phi)
{

  G=length(meth_vec);
  sum=0;
  for(g in 1:G)
	{
		mean=inv.logit(omega_vec%*%alpha[,g]);
		#cat(lambda,"\n")
		sum=sum - dbeta(meth_vec[g],shape1=mean*phi[g],shape2=(1-mean)*phi[g],log=TRUE) ;
		# cat(sum,"\n");
	}

	return(sum);
}


scale=1; N=dim(meth)[1]; G=dim(meth)[2];

K=6;

####  The starting value of omega (topic prop weights)  ####################

# Use a preset seed so the example is reproducable.
require("setRNG")
old.seed <- setRNG(list(kind="Mersenne-Twister", normal.kind="Inversion",
seed=54321))

# omega0=matrix(rdirichlet(N,c(scale/K,scale/K,scale/K,scale/K)), nrow=N);
omega0=omega_initial;
alpha0=matrix(rnorm(K*G,0,1),nrow=(K)); ### the matrix of fixed effects
phi0=rnorm(G,10,1);
param0=c(matrix(omega0,1,N*K),matrix(alpha0,1,K*G),phi0);


library(betareg)
library(lme4)
alpha=matrix(0,K,G);

library(SQUAREM)

beta_est.probe <- function(param_vec,y)
{
	meth=matrix(y,N,G);
	omega0=matrix(param_vec[1:(N*K)],N,K);

	############  estimation of alpha (effect) and phi (disp) #############

	alpha=matrix(0,K,G);
	phi=array(0,G);
	for(g in 1:G)
	{
		meth_col=meth[,g];
    meth_col[which(meth_col<1e-02)]=runif(length(meth_col<1e-02),0,1e-02);
    meth[,g]=meth_col;
		fit=betareg(meth_col~omega0[,1]+omega0[,2]+omega0[,3]+omega0[,4]+omega0[,5]+omega0[,6]-1);
    # fit=betareg(meth_col~omega0[,1]+omega0[,2]-1)
		alpha[,g]=as.numeric(fit$coef[1]$mean);
		phi[g]=as.numeric(fit$coef[2]);
	}

	omega=matrix(0,N,K);
	for(n in 1:dim(meth)[1])
	{
		omega_vec=omega0[n,];
		meth_vec=meth[n,];
		res=optim(reverse_transform(omega_vec), function(v) omega_loglik_beta(transform(v),meth_vec,alpha,phi) );
		omega[n,]=transform(res$par);
	}

	param_vec_omega=matrix(omega,1,N*K);
	param_vec_alpha=matrix(alpha,1,K*G);
	param_new_vec=c(param_vec_omega,param_vec_alpha,phi);
	param_vec=param_new_vec
	return(param_new_vec)
}


options(warn=-1)
system.time(res <- squarem(p=param0,y=y, fixptfn=beta_est.probe, control=list(maxiter = 10, trace = FALSE)));

omega_final=matrix(res$par[1:(N*K)],N,K);
alpha_final_vec=res$par[-(1:(N*K))];
alpha_final=matrix(alpha_final_vec[1:(K*G)],K,G);
phi_final=tail(alpha_final_vec,G);


windows()
k=K
barplot(t(omega_final),col=2:(k+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
abline(v=17)
title(main=paste("Structure Plot of final estd. topic proportions,k=",K))
axis(1,at=1:57,1:57)

write.table(phi_final,"D:/Matthew_Stephens_Project/Methylation/Methylation Expt 2/phi_true.txt")

```

